---
title: "DESeq2"
author: "Li-TIng, Ku"
date: "6/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tximport)
library(readr)
library(tximportData)
library(DESeq2)
library(biomaRt)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(annotate)
library(ComplexHeatmap)
library(viridis)
library(gridExtra)
```

## Get annotation file

- Human : useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl") 
- Mouse: useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl") 

```{r}
## get annotation file
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl") 
t2g <- getBM(attributes = c("ensembl_transcript_id", "ensembl_gene_id", "external_gene_name"), mart = mart)
t2g <- rename(t2g, target_id = ensembl_transcript_id, ens_gene = ensembl_gene_id, ext_gene = external_gene_name)
t2g <- t2g[1:2]
```

## Create metadata file 

- Get the abundance.tsv file from Kallisto in each sample

In this example, PancK1-PancK4 are K700E and PancD1, PancD2, PancH1 are WildType.

```{r}
meta <- data.frame(condition = c(rep("K700E", 4), rep("WT", 3)), rep = c(1:4,1:3))
meta$name <- c("PancK1", "PancK2", "PancK3", "PancK4", "PancD1", "PancD2", "PancH1")
meta$file <- file.path("/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/Gene_tsvfile", meta$name, "abundance.tsv")
```

## PCA plot 

- To see if samples cluster will

```{r}
PCAdata <- tximport(meta$file, type = "kallisto",tx2gene = t2g, ignoreTxVersion=T, ignoreAfterBar = T)$counts
colnames(PCAdata) <- meta$name
res_pca <- prcomp(t(PCAdata))
#fviz_eig(res_pca)
pca <- as.data.frame(res_pca$x)
pca$name <- meta$name
pca$condition <- meta$condition

#pdf("/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/PCA_DGE.pdf", width = 9, height = 8)
ggplot(pca, aes(x = PC1, y = PC2, label = name, color = condition)) + 
  scale_color_manual(values = c("#00798C", "#D1495B", "darkgoldenrod", "#005900", "purple4")) +
  scale_fill_manual(values = c("#00798C", "#D1495B", "darkgoldenrod", "#005900", "purple4")) +
  ggforce::geom_mark_ellipse(aes(fill = condition, group = condition, label = condition), 
                             alpha = 0.1, expand = unit(10, "mm"), lty = "dotted") +
  geom_point() + theme_bw() +
  labs(color = "Sample", fill = "Sample")
#dev.off()
```

## Read in the data table before doing DGE analysis

- We also use MA plot and QC plot to see the overall up and down differential genes and dispersion.

```{r}
kallisto <- tximport(meta$file, type = "kallisto", tx2gene = t2g, ignoreTxVersion=T, ignoreAfterBar = T)
dds <- DESeq(DESeqDataSetFromTximport(kallisto, colData = meta, design = ~ condition))
resultsNames(dds)
# Change the comparison
res <- results(dds, contrast=c('condition','K700E','WT'))
res
#QC plots
#pdf("/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/MA_DGE.pdf", width = 9, height = 8)
plotMA(res,main = "DEGs in K700E vs. WT")
#dev.off()

#pdf("/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/Dispersion_DGE.pdf", width = 9, height = 8)
plotDispEsts(dds, main = "Dispersion of K700E vs WT", legend = F)
#dev.off()
```

## DGE analysis and create the DGE table for down stream anaylses

- Using DESeq2 for normalized counts and annotated by gene name, position.etc.

- In this analysis, I filter out the adjusted p-value with NA, according to this expaination.

http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#indfilt


```{r}
# Create DGE tables
comp <- as.data.frame(res)
exprsn <- as.data.frame(kallisto$abundance)
norm_counts <- counts(dds, normalized = T)
exprsn[,1:7] <- norm_counts
colnames(exprsn) <- meta$name
exprsn <- cbind(exprsn, comp)
exprsn$baseMean <- NULL
#find gene symbols
attributes <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol","start_position", "end_position","chromosome_name", "band"),
                    filters = "ensembl_gene_id", values = row.names(exprsn),
                    mart = mart)
attributes$cyto_band <- paste(attributes$chromosome_name, attributes$band)
exprsn$ENSG <- row.names(exprsn)
exprsn <- dplyr::left_join(exprsn, attributes, by = c("ENSG" = "ensembl_gene_id"))
exprsn <- exprsn[!duplicated(exprsn),]
# resons for filter padj http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#indfilt
exprsn <- exprsn[-which(is.na(exprsn$padj)),]
exprsn <- exprsn[-which(exprsn$hgnc_symbol == ""),]
#write.csv(exprsn, "/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/DESeq2_Panc.csv")
```

## Volcano Plot

- In order to plot the volvano plot, here, I use the cutoff for |log2FoldChange| $\geq$ 2 and adjusted p-value < 0.05
 

```{r}
exprsn$colorgroup <- 0
exprsn[which(abs(exprsn$log2FoldChange) >= 2 & exprsn$padj < 0.05),]$colorgroup <- "group1"
exprsn[which(abs(exprsn$log2FoldChange) >= 2 & exprsn$padj >= 0.05),]$colorgroup <- "group2"
exprsn[which(abs(exprsn$log2FoldChange) < 2 & exprsn$padj < 0.05),]$colorgroup <- "group3"
exprsn[which(abs(exprsn$log2FoldChange) < 2 & exprsn$padj >= 0.05),]$colorgroup <- "group4"

volc <- ggplot(exprsn, aes(x = log2FoldChange, y = -log10(padj), label = hgnc_symbol, color = colorgroup)) +
  geom_point(alpha = 0.5) + 
  geom_hline(yintercept = -log10(0.05), lty = "dashed") +
  geom_vline(xintercept = c(-2,2), lty = "dashed") +
  scale_color_manual(values = c("darkred","darkblue",  "darkgreen", "grey")) +
  theme_classic() +
  theme(legend.position = "none") +
  xlab(expression("WT ← " ~ log[2](FC) ~" → K700E")) + ylab(expression(-log[10]("P-value"))) +
  xlim(-10, 10) +
  geom_text_repel(data = subset(exprsn, abs(exprsn$log2FoldChange) >= 2 & exprsn$padj <= 0.05),
                  box.padding = 0.35,
                  point.padding = 0.5,
                  segment.color = 'grey50',
                  max.overlaps = 40,
                  size = 2) +
  ggtitle("Volcano Plot of Panc0504 DGE K700E vs WT")

#pdf("/Users/litingku/Desktop/Lab/Combine Project SF3b1 RBM10/DGE/Panc 0504/volcplot_DGE.pdf", width = 9, height = 8)
volc
#dev.off()
```
